---
title: "Kōura and their livable lake habitats"
author: "Olivier V. Raven"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    code-fold: true
    code-tools: true
    code-summary: "View Code"
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    embed-resources: true
    smooth-scroll: true
fig-width: 8
fig-height: 6
execute:
  warning: false
  message: false
  cache: true
---

# Livable habitat thresholds
```{r}
#| label: thresholds

# Change these two values to redefine what counts as "livable habitat" for kōura throughout the whole analysis (AEME casts, buoy data, depth ranges, lake maps, trend analysis — every downstream calculation reads from these two variables).

DO_THRESHOLD   <- 5   # mg/L — dissolved oxygen must be >= this value
TEMP_THRESHOLD <- 21  # °C   — temperature must be <= this value
```

# Load packages
```{r}
#| label: setup
#| include: false

# Resolve paths relative to this file — works during render and interactive RStudio sessions
this_file <- knitr::current_input(dir = TRUE)
if (!is.null(this_file) && nzchar(this_file)) {
  base_dir <- normalizePath(dirname(this_file))
} else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  base_dir <- normalizePath(dirname(rstudioapi::getActiveDocumentContext()$path))
} else {
  base_dir <- getwd()
}
knitr::opts_knit$set(root.dir = base_dir)
setwd(base_dir)

options(repos = c(CRAN = "https://cloud.r-project.org"))

Sys.setenv(LERNZMP_API = "lernzmp_lakes")

packages <- c("nzffdr", "sf", "broom", "rLakeAnalyzer", "patchwork", "gganimate", "terra", "pak", "tmap", "tidyverse", "readxl", "ggspatial", "cowplot", "ggrepel", "maptiles", "tidyterra", "osmdata")

quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = TRUE)))
  }
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)
}

invisible(lapply(packages, quiet_load))

if (!requireNamespace("AEME", quietly = TRUE)) {
  pak::pak("limnotrack/AEME")
}
suppressPackageStartupMessages(library(AEME))

if (!requireNamespace("aemetools", quietly = TRUE)) {
  pak::pak("limnotrack/aemetools@dev")
}
suppressPackageStartupMessages(library(aemetools))

if (!requireNamespace("bathytools", quietly = TRUE)) {
  pak::pak("limnotrack/bathytools")
}
suppressPackageStartupMessages(library(bathytools))

```

# Set paths and load the bathymetry data
```{r}
#| label: setup-paths

raw_data_dir <- "data/raw"
der_data_dir <- "data/derived"
out_dir      <- "outputs"

dir.create(der_data_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out_dir,      showWarnings = FALSE, recursive = TRUE)

# Bathymetry is now fetched live from the Limnotrack API for all 12 lakes (see "DATA PREP for maps" section) rather than the local data/raw/bathy_all_lakes/*.tif files.

base_theme_bw <- theme_classic() +
  theme(text     = element_text(family = "Arial", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3))
theme_set(base_theme_bw)

if (knitr::is_latex_output()) {
  theme_update(
    plot.background  = element_rect(fill = "#FFFCF0", colour = NA),
    panel.background = element_rect(fill = "#FFFCF0", colour = NA)
  )
}

```

# Study lakes overview map
```{r}
#| label: fig-study-lakes-map
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| out-width: "100%"
#| fig-cap: "Overview of the 12 Rotorua Te Arawa lakes coloured by Trophic Level Index (TLI, most recent year available from LAWA). Inset shows the location of the study region within Aotearoa New Zealand."

gis_base <- "data/raw/gis_layers"

# Read and validate spatial layers
all_lakes <- sf::st_read(
  file.path(gis_base, "Lakes/12_lakes.gpkg"), quiet = TRUE
) |> sf::st_make_valid() |> sf::st_transform(2193)

nz_land <- sf::st_read(
  file.path(gis_base, "Coastline/nz-coastlines-and-islands-polygons-topo-150k.shp"),
  quiet = TRUE
) |> sf::st_make_valid() |> sf::st_transform(2193)

# Use "name" column (macronated); fall back to "name_ascii", then first non-NA column
non_geom <- names(all_lakes)[!sapply(all_lakes, inherits, what = "sfc")]
name_col <- if ("name" %in% non_geom && any(!is.na(all_lakes[["name"]]))) {
  "name"
} else if ("name_ascii" %in% non_geom) {
  "name_ascii"
} else {
  candidates <- grep("^name", non_geom, ignore.case = TRUE, value = TRUE)
  candidates <- candidates[!sapply(candidates, \(col) all(is.na(all_lakes[[col]])))]
  candidates[1]
}
all_lakes <- dplyr::rename(all_lakes, lake_name = dplyr::all_of(name_col))
all_lakes <- dplyr::mutate(all_lakes,
  # Remove "Lake " prefix
  lake_name = sub("^Lake\\s+", "", lake_name),
  # Keep only the first name — strip " / …", " (…)", " & …" and trailing whitespace
  lake_name = trimws(sub("\\s*[/(\\&].*$", "", lake_name)),
  # Correct missing macrons
  lake_name = dplyr::case_when(
    grepl("^[Oo]kataina",  lake_name)                  ~ "Ōkataina",
    grepl("^[Oo]karo$",    lake_name)                  ~ "Ōkaro",
    grepl("^[Oo]kareka",   lake_name)                  ~ "Ōkareka",
    grepl("^Rotoma$",      lake_name, ignore.case = TRUE) ~ "Rotomā",
    grepl("^Rotokakahi",   lake_name, ignore.case = TRUE) ~ "Rotokākahi",
    TRUE ~ lake_name
  )
)

# ── Trophic state per lake (sourced from LAWA / BOPRC; matches tbl-bop-lake-overview) ──
trophic_lookup <- tibble::tribble(
  ~lake_name,       ~trophic_state,
  "Rotorua",        "Eutrophic",
  "Rotoiti",        "Mesotrophic",
  "Rotoehu",        "Eutrophic",
  "Rotomā",         "Oligotrophic",
  "Ōkataina",       "Oligotrophic",
  "Ōkareka",        "Mesotrophic",
  "Tikitapu",       "Oligotrophic",
  "Rotokākahi",     "Mesotrophic",
  "Tarawera",       "Oligotrophic",
  "Rotomahana",     "Mesotrophic",
  "Ōkaro",          "Eutrophic",
  "Rerewhakaaitu",  "Eutrophic"
)

trophic_colors <- c(
  #"Microtrophic"  = "#3a9dba",
  "Oligotrophic"  = "#7ab648",
  "Mesotrophic"   = "#e9ec0c",
  "Eutrophic"     = "#e26715"
  #"Super-trophic" = "#c0392b"
  )

trophic_lookup <- trophic_lookup |>
  dplyr::mutate(
    trophic_state = factor(trophic_state, levels = names(trophic_colors))
  )

all_lakes <- all_lakes |>
  dplyr::left_join(trophic_lookup, by = "lake_name")

# Map extent
pad      <- 3000
lake_bb  <- sf::st_bbox(all_lakes)
xlim_map <- c(lake_bb["xmin"] - pad, lake_bb["xmax"] + pad)
ylim_map <- c(lake_bb["ymin"] - pad, lake_bb["ymax"] + pad)

# ── Base tile layer: download once per provider, cache as local TIFF ──────────
.provider   <- "Esri.WorldShadedRelief"
.zoom       <- 11
tiles_cache <- file.path(out_dir, paste0("basemap_", gsub("\\.", "_", .provider), ".tif"))
if (!file.exists(tiles_cache)) {
  aoi <- sf::st_as_sfc(lake_bb) |> sf::st_buffer(pad + 5000)
  tiles_raw <- maptiles::get_tiles(aoi, provider = .provider, zoom = .zoom, crop = TRUE)
  terra::writeRaster(tiles_raw, tiles_cache, overwrite = TRUE)
}
basemap <- terra::project(terra::rast(tiles_cache), "EPSG:2193")

# ── Roads: load from cache if available, skip otherwise ──────────────────────
roads_cache <- file.path(out_dir, "roads_osm.rds")
roads <- if (file.exists(roads_cache)) readRDS(roads_cache) else NULL

# ── Label points: extract as plain x/y so ggrepel works without coord_sf stat ─
lake_pts    <- suppressWarnings(sf::st_point_on_surface(all_lakes))
lake_coords <- data.frame(
  lake_name = lake_pts$lake_name,
  x         = sf::st_coordinates(lake_pts)[, 1],
  y         = sf::st_coordinates(lake_pts)[, 2]
)

# ── Main map ──────────────────────────────────────────────────────────────────
p_main <- ggplot() +
  tidyterra::geom_spatraster_rgb(data = basemap) +
  { if (!is.null(roads)) geom_sf(data = roads, color = "#c07030", linewidth = 0.35, alpha = 0.85) } +
  geom_sf(data = all_lakes, aes(fill = trophic_state), color = "#2a2a2a", linewidth = 0.4, alpha = 0.88) +
  scale_fill_manual(
    values   = trophic_colors,
    name     = "Trophic state",
    drop     = FALSE,
    na.value = "grey70",
    guide    = guide_legend(
      title.position = "top",
      keywidth       = unit(0.7, "cm"),
      keyheight      = unit(0.5, "cm"),
      override.aes   = list(alpha = 1, color = NA)
    )
  ) +
  ggrepel::geom_label_repel(
    data             = lake_coords,
    aes(x = x, y = y, label = lake_name),
    size             = 3,
    fontface         = "bold",
    fill             = alpha("white", 0.88),
    color            = "black",
    label.size       = 0.12,
    label.padding    = unit(0.15, "lines"),
    label.r          = unit(0.08, "lines"),
    segment.color    = "black",
    segment.size     = 0.3,
    min.segment.length = 0.2,
    box.padding      = 0.45,
    force            = 3.5,
    max.overlaps     = Inf,
    seed             = 42
  ) +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.28, text_cex = 0.8) +
  ggspatial::annotation_north_arrow(
    location = "tr",
    style    = ggspatial::north_arrow_orienteering(text_size = 7),
    height   = unit(0.9, "cm"), width = unit(0.9, "cm")
  ) +
  coord_sf(xlim = xlim_map, ylim = ylim_map, crs = sf::st_crs(all_lakes),
           expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_void() +
  theme(
    panel.border      = element_rect(fill = NA, color = "black", linewidth = 0.5),
    legend.position   = "none",
    legend.background = element_rect(fill = alpha("white", 0.85), color = "grey60", linewidth = 0.3),
    legend.title      = element_text(size = 9, face = "bold"),
    legend.text       = element_text(size = 8)
  )

# ── Inset NZ map ─────────────────────────────────────────────────────────────
study_box <- sf::st_as_sfc(lake_bb) |> sf::st_buffer(20000)

p_inset <- ggplot() +
  geom_sf(data = nz_land,   fill = "#d4cdb5", color = "#6a5a3a", linewidth = 0.15) +
  geom_sf(data = study_box, fill = alpha("#d62728", 0.25),
          color = "#d62728", linewidth = 0.8) +
  coord_sf(xlim = c(1100000, 2100000), ylim = c(4750000, 6200000),
           crs = sf::st_crs(nz_land), expand = FALSE) +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "#c0d8ec"),
    panel.border     = element_rect(fill = NA, color = "black", linewidth = 0.5)
  )

# ── Combine ───────────────────────────────────────────────────────────────────
# Extract legend from a temporary copy of p_main that has the legend enabled,
# then draw it above the inset in the bottom-right corner.
p_legend_source <- p_main +
  theme(legend.position = c(0.5, 0.5))  # position doesn't matter — just enables it
p_legend <- cowplot::get_legend(p_legend_source)

p_combined <- cowplot::ggdraw(p_main) +
  cowplot::draw_plot(p_inset,  x = 0.68, y = 0.03, width = 0.30, height = 0.33) +
  cowplot::draw_plot(p_legend, x = 0.68, y = 0.36, width = 0.30, height = 0.20)

ggsave(
  file.path(out_dir, "fig-study-lakes-map.png"),
  p_combined, width = 8, height = 8, dpi = 300, bg = "white"
)

p_combined
```

# Kōura and catfish distribution maps (NZFFD)
```{r}
#| label: nz-distribution-maps
#| fig-width: 12
#| fig-height: 9

nzffdms <- read_csv("data/raw/nzffdms.csv", show_col_types = FALSE)
nzffd_clean    <- nzffdr_clean(nzffdms)
nzffd_clean    <- nzffdr_add_dates(nzffd_clean)
nzffd_enhanced <- nzffdr_taxon_threat(nzffd_clean)

# Valid NZTM coordinates only
nzffd_valid <- nzffd_enhanced |>
  filter(!is.na(eastingNZTM), !is.na(northingNZTM))

# Kōura — unique survey locations
nzffd_koura <- nzffd_valid |>
  filter(str_detect(tolower(taxonCommonName), "koura|kōura")) |>
  distinct(eastingNZTM, northingNZTM, .keep_all = TRUE)

# Catfish — unique survey locations
nzffd_catfish <- nzffd_valid |>
  filter(str_detect(tolower(taxonCommonName), "catfish")) |>
  distinct(eastingNZTM, northingNZTM, .keep_all = TRUE)

# Convert to sf (NZTM EPSG:2193) then reproject to WGS84 to match nzffdr_nzmap and coord_sf limits
nzffd_koura_sf   <- st_as_sf(nzffd_koura,   coords = c("eastingNZTM", "northingNZTM"), crs = 2193) |>
  sf::st_transform(4326)
nzffd_catfish_sf <- st_as_sf(nzffd_catfish, coords = c("eastingNZTM", "northingNZTM"), crs = 2193) |>
  sf::st_transform(4326)

nz_map <- nzffdr_nzmap

map_theme <- theme_void() +
  theme(
    plot.title    = element_text(face = "bold", size = 13, hjust = 0.5, margin = margin(b = 3)),
    plot.subtitle = element_text(size = 9, hjust = 0.5, colour = "grey40", margin = margin(b = 6)),
    plot.margin   = margin(8, 8, 8, 8)
  )

# Crop to NZ main islands — excludes Chatham Islands (far east, ~176°W)
nz_main_xlim <- c(165.5, 178.8)
nz_main_ylim <- c(-47.8, -33.8)

koura_map <- ggplot() +
  geom_sf(data = nz_map, fill = "grey92", colour = "grey60", linewidth = 0.3) +
  geom_sf(data = nzffd_koura_sf, colour = "#2171b5", size = 0.7, alpha = 0.5, shape = 16) +
  coord_sf(xlim = nz_main_xlim, ylim = nz_main_ylim, crs = 4326) +
  labs(
    title    = "Kōura",
    subtitle = paste0(format(nrow(nzffd_koura), big.mark = ","), " survey locations")
  ) +
  map_theme

catfish_map <- ggplot() +
  geom_sf(data = nz_map, fill = "grey92", colour = "grey60", linewidth = 0.3) +
  geom_sf(data = nzffd_catfish_sf, colour = "#d73027", size = 0.7, alpha = 0.5, shape = 16) +
  coord_sf(xlim = nz_main_xlim, ylim = nz_main_ylim, crs = 4326) +
  labs(
    title    = "Brown bullhead catfish",
    subtitle = paste0(format(nrow(nzffd_catfish), big.mark = ","), " survey locations")
  ) +
  map_theme

nz_distribution <- koura_map + catfish_map +
  plot_annotation(
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey40")
    )
  )

nz_distribution

ggsave(file.path(out_dir, "fig-nz-distribution-maps.png"), nz_distribution,
       width = 12, height = 9, dpi = 300)
```

# Run LERNZmp (AEME)
```{r}
#| label: aeme-load-or-run

hypsograph_path    <- file.path(der_data_dir, "hypsograph_data.csv")
water_quality_path <- file.path(der_data_dir, "water_quality_data.csv")

if (file.exists(hypsograph_path) && file.exists(water_quality_path)) {

  # Derived files already exist — skip the slow model loop
  hypsograph_data    <- read_csv(hypsograph_path,    show_col_types = FALSE)
  water_quality_data <- read_csv(water_quality_path, show_col_types = FALSE)

} else {

  # All 12 Rotorua Te Arawa lakes (Bay of Plenty Regional Council programme)
  lid_files <- c("LID11133", "LID14290", "LID15325", "LID40102", "LID40188", "LID54730",
                 "LID15312", "LID15621", "LID40071", "LID54731", "LID54732", "LID54733")

  metadata <- read.csv("data/raw/lernzmp/LERNZmp_lake_metadata.csv") |>
    dplyr::filter(aeme_file %in% lid_files)

  model <- c("glm_aed", "gotm_wet")
  path  <- "aeme"

  hypsograph_data    <- tibble()
  water_quality_data <- tibble()

  for (i in seq_along(lid_files)) {
    lid     <- lid_files[i]
    lake_id <- as.integer(sub("^LID", "", lid))

    result <- tryCatch({
      aeme <- aemetools::get_aeme(id = lake_id, api_key = "lernzmp_lakes")
      if (lid == "LID40188") {
        shoreline    <- aemetools::get_lake_shape(id = 40188, api_key = "lernzmp_lakes")
        contours     <- aemetools::lt_fetch(
          "lake_contours",
          filter = aemetools::lt_filter(lernzmp_id == "LID40188")
        ) |>
          dplyr::mutate(depth = depth_m)
        bathy_raster <- rasterise_bathy(shoreline = shoreline, contours = contours,
                                        res = 8, crs = 2193)
        lake_elev    <- max(contours$elevation_m)
        upd_hyps     <- bathy_to_hypso(bathy_raster) |>
          dplyr::mutate(elev = lake_elev + depth) |>
          extrap_hyps(z_range = 0.05, ext_elev = 3)
        aeme <- add_hypsograph(aeme, hypsograph = upd_hyps)
      }

      inp           <- input(aeme)
      lake_metadata <- metadata |> dplyr::filter(aeme_file == lid)

      hyps <- inp$hypsograph |>
        dplyr::filter(depth <= 0) |>
        mutate(
          LID               = lid,
          lake_name         = lake_metadata$Name,
          region            = lake_metadata$Region,
          geomorphic_type   = lake_metadata$Geomorphic.type,
          max_depth         = lake_metadata$Depth,
          total_area        = lake_metadata$Area,
          depth_measurement = lake_metadata$Depth.measurement,
          data_quality      = lake_metadata$Data,
          area_ha           = area / 10000
        )

      lake_observations <- aeme@observations[["lake"]] |>
        dplyr::select(Date, depth_mid, var_aeme, value) |>
        mutate(
          lake_name         = lake_metadata$Name,
          region            = lake_metadata$Region,
          geomorphic_type   = lake_metadata$Geomorphic.type,
          max_depth         = lake_metadata$Depth,
          depth_measurement = lake_metadata$Depth.measurement,
          data_quality      = lake_metadata$Data
        )

      message("AEME data fetched for ", lid, " (", lake_metadata$Name, ")")
      list(hyps = hyps, obs = lake_observations)

    }, error = function(e) {
      message("Skipping ", lid, " — no AEME model/data available: ", conditionMessage(e))
      NULL
    })

    if (!is.null(result)) {
      hypsograph_data    <- bind_rows(hypsograph_data,    result$hyps)
      water_quality_data <- bind_rows(water_quality_data, result$obs)
    }
  }

  write_csv(hypsograph_data,    hypsograph_path)
  write_csv(water_quality_data, water_quality_path)
}

```

# DATA PREP (functions)
```{r}
#---- Global settings / helpers ----
# All 12 Rotorua Te Arawa lakes — fixed display/facet order
lake_order <- c("Rotorua", "Rotoiti", "Rotoehu", "Rotomā",
                "Ōkataina", "Ōkāreka", "Tikitapu", "Rotokākahi",
                "Tarawera", "Rotomahana", "Ōkaro", "Rerewhakaaitu")
season_order <- c("Summer", "Autumn", "Winter", "Spring")

#Fetch buoy water-quality profiler data from the Limnotrack API
get_lake_wqprofiler <- function(lake = "rotorua", type = "pro",
                                api_url = "https://api.limnotrack.com",
                                api_key = "lernzmp_lakes") {

  res <- aemetools::api_request(api_url = api_url, api_key = api_key,
                                endpoint = "get_lake_wqprofiler",
                                query = list(lake = lake, type = type))

  if (httr2::resp_status(res) != 200) {
    cli::cli_abort("API request failed with status: ", httr2::resp_status(res))
  }

  res |>
    httr2::resp_body_string() |>
    jsonlite::fromJSON()
}

recode_lake_names <- function(x) {
dplyr::recode(x,
  "Rotoma" = "Rotomā", "Okareka" = "Ōkāreka", "Okaro" = "Ōkaro",
  "Okataina" = "Ōkataina", "Rotokakahi" = "Rotokākahi"
)
}

add_season_nz <- function(df, date_col = "Date") {
df |>
mutate(
!!date_col := as.Date(.data[[date_col]]),
Day = lubridate::day(.data[[date_col]]),
# Month label is locale-dependent (used for display only) — Season below uses
# the numeric month so it doesn't break under non-English R session locales.
Month = lubridate::month(.data[[date_col]], label = TRUE, abbr = TRUE),
Month_num = lubridate::month(.data[[date_col]]),
Year = lubridate::year(.data[[date_col]]),
Season = case_when(
Month_num %in% c(12, 1, 2) ~ "Summer",
Month_num %in% c(3, 4, 5)  ~ "Autumn",
Month_num %in% c(6, 7, 8)  ~ "Winter",
Month_num %in% c(9, 10, 11) ~ "Spring",
TRUE ~ NA_character_
),
Season = factor(Season, levels = season_order),
# Southern Hemisphere summer spans the New Year (Dec-Jan-Feb), so a plain calendar
# Year splits one continuous summer into two groups. Convention used here: December
# rolls forward into the FOLLOWING year's summer label, so "Summer 2021" = Dec 2020 +
# Jan 2021 + Feb 2021 — matching how NZ/climate-science usage names a summer after the
# January/February it mostly falls in (i.e. "summer 2020/21" -> Season_Year 2021).
# Autumn/Winter/Spring don't cross the New Year, so Season_Year == Year for those.
Season_Year = if_else(Month_num == 12, Year + 1L, Year)
) |>
dplyr::select(-Month_num)
}

#Compute livable habitat per cast/profile using the same logic for both sources
#(Non-livable area summed across depth layers; livable = surface area - non-livable)

calc_livable_habitat_summary <- function(df,
source_label,
lake_col = "Lake",
date_col = "Date",
cast_col = "Cast_ID",
depth_col = "Depth",
do_col = "DO",
temp_col = "Temp",
area0_col = "Area_depth_0",
arealayer_col = "Area_at_Depth",
do_threshold = DO_THRESHOLD,
temp_threshold = TEMP_THRESHOLD) {
df |>
mutate(
Source = source_label,
Lake = recode_lake_names(.data[[lake_col]]),
Lake = stringr::str_replace(Lake, "\\s*\\(.*\\)", "")) |>
add_season_nz(date_col = date_col) |>
mutate(
Livable_Habitat = .data[[do_col]] >= do_threshold & .data[[temp_col]] <= temp_threshold) |>
group_by(Lake, .data[[date_col]], .data[[cast_col]]) |>
summarise(
Area_depth_0 = first(.data[[area0_col]]),
Day = first(Day),
Month = first(Month),
Year = first(Year),
Season = first(Season),
Non_livable_area = sum(.data[[arealayer_col]][!Livable_Habitat], na.rm = TRUE),
Sum_Livable_Habitat_Area = Area_depth_0 - Non_livable_area,
Livable_Habitat_Percentage = (Sum_Livable_Habitat_Area / Area_depth_0) * 100,
.groups = "drop"
) |>
mutate(
Lake = factor(Lake, levels = lake_order),
Season = factor(Season, levels = season_order)
)
}

calc_depth_ranges <- function(df,
                             source_label,
                             lake_col = "Lake",
                             date_col = "Date",
                             cast_col = "Cast_ID",
                             depth_col = "Depth",
                             livable_col = "Livable_Habitat") {

  df1 <- df |>
    mutate(
      Lake = recode_lake_names(.data[[lake_col]]),
      Lake = stringr::str_replace(Lake, "\\s*\\(.*\\)", ""),
      Lake = factor(Lake, levels = lake_order)
    ) |>
    add_season_nz(date_col = date_col) |>
    group_by(Lake, Year, Season, .data[[date_col]], .data[[cast_col]]) |>
    summarise(
      cast_shallowest = min(.data[[depth_col]], na.rm = TRUE),
      cast_deepest    = max(.data[[depth_col]], na.rm = TRUE),
      shallowest_livable = ifelse(any(.data[[livable_col]], na.rm = TRUE),
        min(.data[[depth_col]][.data[[livable_col]]], na.rm = TRUE), NA_real_),
      deepest_livable = ifelse(any(.data[[livable_col]], na.rm = TRUE),
        max(.data[[depth_col]][.data[[livable_col]]], na.rm = TRUE), NA_real_),
      .groups = "drop"
    )

  df1 |>
    group_by(Lake, Year, Season) |>
    summarise(
      Date = as.Date(first(.data[[date_col]])),
      cast_shallowest = min(cast_shallowest, na.rm = TRUE),
      cast_deepest    = max(cast_deepest, na.rm = TRUE),
      shallowest_livable = mean(shallowest_livable, na.rm = TRUE),
      deepest_livable    = mean(deepest_livable, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      Source = source_label,
      Lake   = factor(Lake, levels = lake_order),
      Season = factor(Season, levels = season_order)
    )
}


```


# DATA PREP (AEME profiles + hypsograph join)
```{r}
hypsograph_data <- read_csv("data/derived/hypsograph_data.csv", show_col_types = FALSE)
water_quality_data <- read_csv("data/derived/water_quality_data.csv", show_col_types = FALSE)

#Keep DO and temperature; clean lake names

DO_Tem_data <- water_quality_data |>
filter(var_aeme %in% c("CHM_oxy", "HYD_temp")) |>
mutate(lake_name = stringr::str_replace(lake_name, "\\s*\\(.*\\)", ""))

#Optional check plot

ggplot(DO_Tem_data, aes(value, depth_mid, colour = Date)) +
geom_line() +
facet_wrap(var_aeme ~ lake_name, scales = "free") +
scale_y_reverse()

#Wide DO/Temp table

DO_Tem_data1 <- DO_Tem_data |>
dplyr::select(Date, depth_mid, var_aeme, value, lake_name) |>
# The live API can return duplicate readings for the same Date/depth/lake/variable
# (the old local .rds snapshots were already deduplicated) — average them so
# pivot_wider gets one value per cell instead of a list-col.
tidyr::pivot_wider(names_from = var_aeme, values_from = value, values_fn = mean) |>
rename(
Lake = lake_name,
Depth = depth_mid,
DO = CHM_oxy,
Temp = HYD_temp
) |>
dplyr::select(Lake, Date, Depth, DO, Temp)

#Cast IDs per lake-date profile

DO_Tem_data1 <- DO_Tem_data1 |>
group_by(Lake, Date) |>
mutate(Cast_ID = dplyr::cur_group_id()) |>
ungroup()

#Hypsograph processing: area per depth layer

hypsograph_data1 <- hypsograph_data |>
filter(depth == floor(depth)) |>
mutate(
depth = abs(depth),
Area_at_Depth = abs(area_ha - lag(area_ha, default = first(area_ha)))
) |>
filter(!is.na(area_ha)) |>
arrange(lake_name, depth) |>
group_by(lake_name) |>
mutate(
Area_at_Depth = if_else(depth == 0, 0, Area_at_Depth),
Area_depth_0 = first(area_ha)
) |>
ungroup() |>
dplyr::select(
Lake = lake_name,
Depth = depth,
Area = area_ha,
Area_at_Depth,
Area_depth_0,
Total_area = total_area
) |>
mutate(Lake = stringr::str_replace(Lake, "\\s*\\(.*\\)", ""))

```

# Check hypsographs — verify depth profiles before building maps
```{r}
#| label: fig-hypsograph-check
#| fig-width: 12
#| fig-height: 10

hypsograph_check_data <- hypsograph_data1 |>
  mutate(Lake = recode_lake_names(Lake), Lake = factor(Lake, levels = lake_order))

# Flag lakes where the shallowest measured depth is > 1 m (missing littoral zone)
shallow_check <- hypsograph_check_data |>
  group_by(Lake) |>
  summarise(
    min_depth    = min(Depth, na.rm = TRUE),
    max_depth    = max(Depth, na.rm = TRUE),
    n_depths     = n(),
    area_at_0    = Area[Depth == min(Depth)][1],
    .groups = "drop"
  ) |>
  mutate(
    shallow_ok = min_depth <= 1,
    coverage   = round(n_depths / (max_depth + 1), 2)
  )

print(shallow_check)

# Depths where area does not change between consecutive steps
flat_steps <- hypsograph_check_data |>
  arrange(Lake, Depth) |>
  group_by(Lake) |>
  mutate(area_change = Area - lag(Area)) |>
  filter(!is.na(area_change), area_change == 0) |>
  select(Lake, Depth, Area) |>
  ungroup()

if (nrow(flat_steps) > 0) {
  message("Depth steps with no area change:")
  print(flat_steps)
} else {
  message("No flat steps found — area changes at every depth.")
}

# Check depth resolution in the shallow zone (0–10 m)
# A gap > 1 m here causes the visible straight-line drop in the hypsograph
shallow_gaps <- hypsograph_check_data |>
  arrange(Lake, Depth) |>
  filter(Depth <= 10) |>
  group_by(Lake) |>
  mutate(
    depth_gap   = Depth - lag(Depth),
    area_change = Area  - lag(Area)
  ) |>
  filter(!is.na(depth_gap), depth_gap > 1) |>
  select(Lake, from_depth = Depth, depth_gap, Area, area_change) |>
  ungroup()

if (nrow(shallow_gaps) > 0) {
  message("Shallow-zone gaps > 1 m (potential straight-line artefact in hypsograph):")
  print(shallow_gaps)
} else {
  message("Shallow zone (0–10 m) measured at ≤ 1 m intervals for all lakes.")
}

# Lakes with potential shallow-zone problems
problem_lakes <- shallow_check |> filter(!shallow_ok) |> pull(Lake)
if (length(problem_lakes) > 0) {
  message("Lakes missing shallow (<= 1 m) bathymetry: ", paste(problem_lakes, collapse = ", "))
} else {
  message("All lakes have depth measurements at or near 0 m.")
}

hypsograph_check_plot <- hypsograph_check_data |>
  ggplot(aes(x = Area, y = Depth)) +
  geom_line(colour = "#0065a9", linewidth = 0.8) +
  geom_point(size = 0.8, colour = "#0065a9") +
  # Red point at the shallowest measured depth so problem lakes are immediately visible
  geom_point(
    data = hypsograph_check_data |>
      group_by(Lake) |> slice_min(Depth, n = 1, with_ties = FALSE),
    colour = "red", size = 2.5, shape = 21, fill = "red", stroke = 0.5
  ) +
  scale_y_reverse() +
  facet_wrap(~ Lake, scales = "free", ncol = 3) +
  labs(x = "Area (ha)", y = "Depth (m)",
       title = "Hypsograph check — red dot = shallowest measured depth (should be at or near 0 m)",
       caption = paste0(
         if (length(problem_lakes) > 0)
           paste("Potential shallow-zone gaps:", paste(problem_lakes, collapse = ", "))
         else
           "All lakes measured to near-surface depth."
       ))

hypsograph_check_plot
ggsave(file.path(out_dir, "fig-hypsograph-check.png"), hypsograph_check_plot,
       width = 12, height = 10, dpi = 300)
```

```{r}
#Combine DO/Temp profiles with hypsograph

DO_Tem_area_data <- DO_Tem_data1 |>
left_join(hypsograph_data1, by = c("Lake", "Depth")) |>
filter(!is.na(Area))

#Final standardise

DO_Tem_area_data <- DO_Tem_area_data |>
mutate(
Lake = recode_lake_names(Lake),
Lake = factor(Lake, levels = lake_order)
) |>
add_season_nz("Date")

#Temp and DO plots

DO_Tem_area_data$Season <- factor(DO_Tem_area_data$Season, levels = season_order)

TEMP_plot <- ggplot(DO_Tem_area_data, aes(Temp, Depth, col = Season, group = Cast_ID)) +
geom_line() +
geom_vline(xintercept = TEMP_THRESHOLD, linetype = "dashed") +
scale_y_reverse() +
facet_wrap(~ Lake, scales = "free_y", ncol = 3)

DO_plot <- ggplot(DO_Tem_area_data, aes(DO, Depth, col = Season, group = Cast_ID)) +
geom_line() +
geom_vline(xintercept = DO_THRESHOLD, linetype = "dashed") +
scale_y_reverse() +
facet_wrap(~ Lake, scales = "free_y", ncol = 3)

TEMP_plot / DO_plot

```


# Livable habitat — cast data prep (AEME)
```{r}
#| label: livable-cast-prep

DO_Tem_area_data <- DO_Tem_area_data |>
  mutate(Livable_Habitat = DO >= DO_THRESHOLD & Temp <= TEMP_THRESHOLD)

livable_habitat_sum <- calc_livable_habitat_summary(
  df           = DO_Tem_area_data,
  source_label = "Cast",
  lake_col     = "Lake",  date_col     = "Date",   cast_col     = "Cast_ID",
  depth_col    = "Depth", do_col       = "DO",     temp_col     = "Temp",
  area0_col    = "Area_depth_0",                   arealayer_col = "Area_at_Depth"
)
```

# DATA PREP for maps
```{r}
# All lakes now have bathymetry (fetched via API)
area_data_all_lakes <- DO_Tem_area_data

# First cast per Lake–Year–Month
monthly_casts <- area_data_all_lakes |>
distinct(Lake, Year, Month, Date, Cast_ID) |>
arrange(Lake, Year, Month, Date) |>
group_by(Lake, Year, Month) |>
slice(1) |>
ungroup()

bathy_to_points <- function(r, lake_name) {
df <- as.data.frame(r, xy = TRUE, na.rm = TRUE)
names(df)[3] <- "depth"
df |> mutate(Lake = lake_name)
}

# Bathymetry for all 12 lakes — fetched live from the Limnotrack API (lake shoreline +
# depth contours, rasterised), the same approach used for the Rotoehu hypsograph fix.
# Cached to disk the same way as the AEME CSVs / livable_combined.rds: load straight
# from bathy_rasters.rds if it already exists, otherwise fetch from the API (which has
# turned out to be quite flaky/intermittent) and save so we don't need to hit it again.
# Delete data/derived/bathy_rasters.rds to force a fresh re-fetch from the API.
bathy_rasters_rds <- file.path(der_data_dir, "bathy_rasters.rds")

if (file.exists(bathy_rasters_rds)) {

  bathy_rasters <- readRDS(bathy_rasters_rds) |> lapply(terra::unwrap)

} else {

  bathy_lakes <- tibble::tribble(
    ~lake_id, ~lake_name,
    11133,    "Rotorua",
    54730,    "Rotoiti",
    40188,    "Rotoehu",
    40102,    "Rotomā",
    54731,    "Ōkataina",
    15325,    "Ōkāreka",
    15312,    "Tikitapu",
    15621,    "Rotokākahi",
    54732,    "Tarawera",
    54733,    "Rotomahana",
    14290,    "Ōkaro",
    40071,    "Rerewhakaaitu"
  )

  # lt_filter()'s server-side NSE doesn't reliably see local variables (confirmed: both a
  # plain variable and rlang `!!` injection fail with "object not found", and in the original
  # session it silently fell back to stale/wrong data instead of erroring). Side-step it
  # entirely — fetch every lake's contours once, unfiltered, then filter locally per lake
  # with plain dplyr::filter(), which we know works correctly everywhere else in this script.
  all_contours <- aemetools::lt_fetch("lake_contours", limit = 50000)

  # Lakes with no contour data on the API fall back to these local raster snapshots
  local_bathy_files <- c(
    "54730" = "data/raw/bathy_all_lakes/54730_lake_rotoiti.tif"
  )

  fetch_bathy_raster <- function(lake_id, lake_name) {
    tryCatch({
      shoreline <- aemetools::get_lake_shape(id = lake_id, api_key = "lernzmp_lakes")
      lid_str   <- paste0("LID", lake_id)
      contours  <- all_contours |>
        dplyr::filter(lernzmp_id == lid_str) |>
        dplyr::mutate(depth = depth_m)
      if (nrow(contours) == 0) stop("No contour rows found for ", lid_str)
      # Scale resolution to lake size (~150x150 grid) so small lakes (e.g. Tikitapu, Ōkaro)
      # get crisp detail instead of the blocky look a fixed 50m grid gives them, while huge
      # lakes (e.g. Rotorua) stay capped so interpolation doesn't get unreasonably slow.
      lake_area_m2 <- as.numeric(sf::st_area(shoreline))
      res_dynamic  <- pmin(50, pmax(5, round(sqrt(lake_area_m2) / 150)))
      bathy_raster <- rasterise_bathy(shoreline = shoreline, contours = contours, res = res_dynamic, crs = 2193)
      message("Bathymetry fetched for ", lake_name, " (LID", lake_id, ")")
      terra::wrap(round(abs(bathy_raster)))
    }, error = function(e) {
      local_path <- local_bathy_files[as.character(lake_id)]
      if (!is.na(local_path) && file.exists(local_path)) {
        message("No API contours for ", lake_name, " (LID", lake_id, ") — using local raster fallback")
        return(terra::wrap(round(abs(terra::rast(local_path)))))
      }
      message("Skipping bathymetry for ", lake_name, " (LID", lake_id, "): ", conditionMessage(e))
      NULL
    })
  }

  bathy_rasters <- Map(fetch_bathy_raster, bathy_lakes$lake_id, bathy_lakes$lake_name)
  names(bathy_rasters) <- bathy_lakes$lake_name
  bathy_rasters <- bathy_rasters[!sapply(bathy_rasters, is.null)]

  # Don't cache an obviously incomplete result (e.g. a full API outage where almost
  # everything failed) — that would lock in the bad result for every future run until
  # someone notices and manually deletes the file. Only save if most lakes succeeded.
  if (length(bathy_rasters) >= round(nrow(bathy_lakes) * 0.5)) {
    saveRDS(bathy_rasters, bathy_rasters_rds)
  } else {
    warning("Only ", length(bathy_rasters), "/", nrow(bathy_lakes),
            " lakes got bathymetry this run — not caching (likely API outage). ",
            "Re-run later once the API is back up.")
  }
  bathy_rasters <- lapply(bathy_rasters, terra::unwrap)

}

bathy_points <- purrr::imap(bathy_rasters, bathy_to_points) |> bind_rows()

max_depth_lake <- bathy_points |>
group_by(Lake) |>
summarise(Max_depth = max(depth, na.rm = TRUE), .groups = "drop")

set.seed(1)
bathy_points_sub <- bathy_points |> group_by(Lake) |> slice_sample(n = 20000) |> ungroup()
bathy_grid <- bathy_points |> mutate(Depth_round5 = floor(depth / 5) * 5)

profile_conditions <- area_data_all_lakes |>
select(Lake, Date, Cast_ID, Depth, Year, Season, Livable_Habitat)

bathy_cast <- bathy_points_sub |>
rename(Depth = depth) |>
left_join(profile_conditions, by = c("Lake", "Depth")) |>
mutate(
Lake = factor(Lake, levels = lake_order),
Cast_label = paste(Year, format(Date, "%Y-%m-%d"), Cast_ID, Lake, Season, sep = " | ")
) |>
arrange(Year, Date, Cast_ID, Lake)

```


# Data prep — buoy CTD data
```{r}
#| label: buoy-fetch

buoy_raw_rds <- file.path(der_data_dir, "buoy_raw_data.rds")

if (file.exists(buoy_raw_rds)) {

  CTD_buoy <- readRDS(buoy_raw_rds)

} else {

  # Lake slugs match the API's naming (lowercase, no macrons); Title-cased afterwards so
  # recode_lake_names() can map them to the macron spellings used everywhere else.
  buoy_lake_slugs <- c("rotorua", "rotoiti", "rotoehu", "rotoma", "okareka", "okaro",
                       "tikitapu", "rotokakahi", "rerewhakaaitu", "okataina", "tarawera", "rotomahana")

  fetch_buoy_api <- function(lake_slug) {
    tryCatch({
      df <- get_lake_wqprofiler(lake_slug)
      df$Lake <- stringr::str_to_title(lake_slug)
      message("Buoy data fetched for ", lake_slug, ": ", nrow(df), " rows")
      df
    }, error = function(e) {
      message("No buoy data from API for ", lake_slug, ": ", conditionMessage(e))
      NULL
    })
  }

  CTD_buoy_list <- lapply(buoy_lake_slugs, fetch_buoy_api)
  names(CTD_buoy_list) <- buoy_lake_slugs

  # Fall back to local CSVs for any lake the API didn't return (currently only Ōkaro has these)
  if (is.null(CTD_buoy_list[["okaro"]])) {
    Okaro_CTD_buoy_files <- list.files("data/raw/Okaro_buoy_profiles", pattern = "\\.csv$", full.names = TRUE)
    if (length(Okaro_CTD_buoy_files) > 0) {
      message("Falling back to local CSVs for okaro")
      Okaro_CTD_buoy        <- bind_rows(lapply(Okaro_CTD_buoy_files, read.csv))
      Okaro_CTD_buoy$Lake   <- "Okaro"
      CTD_buoy_list[["okaro"]] <- Okaro_CTD_buoy
    }
  }

  CTD_buoy <- bind_rows(CTD_buoy_list)
  if (nrow(CTD_buoy) == 0) stop("No buoy CTD data could be fetched for any lake")

  saveRDS(CTD_buoy, buoy_raw_rds)

}
```

# Data prep — livable habitat combined (Cast + Buoy CTD)
```{r}
#| label: livable-combined-prep

combined_rds <- file.path(der_data_dir, "livable_combined.rds")

if (file.exists(combined_rds)) {

  saved               <- readRDS(combined_rds)
  livable_habitat_sum <- saved$livable_habitat_sum
  livable_habitat_all <- saved$livable_habitat_all
  depth_ranges_all    <- saved$depth_ranges_all

} else {

  # ---- Cast depth ranges (all lakes) ----
  livable_depth_season <- calc_depth_ranges(
    df           = area_data_all_lakes,
    source_label = "Cast",
    lake_col     = "Lake", date_col  = "Date",  cast_col  = "Cast_ID",
    depth_col    = "Depth", livable_col = "Livable_Habitat"
  ) |>
    left_join(max_depth_lake, by = "Lake") |>
    mutate(Lake = factor(Lake, levels = lake_order))

  # CTD_buoy comes from the independently-cached "buoy-fetch" chunk above —
  # no re-fetching needed here.
  Buoy_data <- CTD_buoy |>
    dplyr::select(Lake, DateTime, DptSns, TmpWtr, DOconc) |>
    rename(Depth = DptSns, Temp = TmpWtr, DO = DOconc) |>
    distinct(Lake, DateTime, .keep_all = TRUE) |>
    mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S")) |>
    tidyr::drop_na(DateTime, Depth) |>
    mutate(
      Date  = as.Date(DateTime),
      Time  = format(DateTime, "%H:%M:%S"),
      Year  = lubridate::year(DateTime),
      # Month label is locale-dependent (display only) — Season uses numeric month instead
      Month = lubridate::month(DateTime, label = TRUE),
      Day   = lubridate::day(DateTime),
      Season = case_when(
        lubridate::month(DateTime) %in% c(12, 1, 2) ~ "Summer",
        lubridate::month(DateTime) %in% c(3, 4, 5)  ~ "Autumn",
        lubridate::month(DateTime) %in% c(6, 7, 8)  ~ "Winter",
        lubridate::month(DateTime) %in% c(9, 10, 11) ~ "Spring",
        TRUE ~ NA_character_
      ),
      Date_Time_Numeric = as.numeric(DateTime),
      Time_of_Day = case_when(
        lubridate::hour(DateTime) >= 6  & lubridate::hour(DateTime) < 12 ~ "Morning",
        lubridate::hour(DateTime) >= 12 & lubridate::hour(DateTime) < 18 ~ "Afternoon",
        lubridate::hour(DateTime) >= 18 & lubridate::hour(DateTime) < 24 ~ "Evening",
        TRUE ~ "Night"
      )
    )

  Buoy_data1 <- Buoy_data |>
    arrange(Lake, Date, Time) |>
    mutate(
      Depth   = round(Depth),
      cast_id = cumsum(Depth < lag(Depth, default = first(Depth))) + 1,
      cast_id = as.factor(cast_id)
    ) |>
    group_by(Lake, Date) |>
    mutate(num_casts = n_distinct(cast_id)) |>
    ungroup() |>
    group_by(Lake, Date, cast_id, Depth) |>
    summarise(
      Temp = mean(Temp, na.rm = TRUE),
      DO   = mean(DO,   na.rm = TRUE),
      across(-c(Temp, DO), dplyr::first),
      .groups = "drop"
    )

  Buoy_data2 <- Buoy_data1 |>
    left_join(hypsograph_data1, by = c("Lake", "Depth")) |>
    mutate(
      Lake = recode_lake_names(Lake),
      Lake = factor(Lake, levels = lake_order),
      Livable_Habitat = DO >= DO_THRESHOLD & Temp <= TEMP_THRESHOLD
    )

  Buoy_livable_habitat <- calc_livable_habitat_summary(
    df           = Buoy_data2 |> rename(Cast_ID = cast_id),
    source_label = "CTD",
    lake_col     = "Lake",  date_col     = "Date",   cast_col     = "Cast_ID",
    depth_col    = "Depth", do_col       = "DO",     temp_col     = "Temp",
    area0_col    = "Area_depth_0",                   arealayer_col = "Area_at_Depth"
  )

  buoy_livable_depth_season <- calc_depth_ranges(
    df           = Buoy_data2 |> rename(Cast_ID = cast_id),
    source_label = "CTD",
    lake_col     = "Lake",  date_col    = "Date",   cast_col    = "Cast_ID",
    depth_col    = "Depth", livable_col = "Livable_Habitat"
  )

  # ---- Combine Cast + Buoy ----
  livable_habitat_all <- bind_rows(
    livable_habitat_sum  |> mutate(Source = "Cast", Cast_ID = as.character(Cast_ID)),
    Buoy_livable_habitat |> mutate(Source = "CTD",  Cast_ID = as.character(Cast_ID))
  ) |>
    mutate(
      Source = factor(Source, levels = c("Cast", "CTD")),
      Lake   = factor(Lake,   levels = lake_order),
      Season = factor(Season, levels = season_order)
    )

  depth_ranges_all <- bind_rows(
    livable_depth_season      |> mutate(Source = "Cast"),
    buoy_livable_depth_season |> mutate(Source = "CTD")
  ) |>
    mutate(
      Source = factor(Source, levels = c("Cast", "CTD")),
      Lake   = factor(Lake,   levels = lake_order),
      Season = factor(Season, levels = season_order)
    )

  saveRDS(
    list(
      livable_habitat_sum = livable_habitat_sum,
      livable_habitat_all = livable_habitat_all,
      depth_ranges_all    = depth_ranges_all
    ),
    combined_rds
  )

}
```

# Historical profiles 1975–1976 (Lake Rotoiti)
```{r}
#| label: historical-rotoiti-1975-1976

# Surface (0 m) and deep (~50 m) values read from annotated numbers in: Devcich (1979) Fig 26: oxygen profiles (g/m³ = mg/L); Fig 30: temperature profiles (°C) Intermediate depths generated by sigmoid interpolation calibrated to the visible curve shapes (sharp thermocline/oxycline ~10–15 m in summer, near-uniform in winter).  Values marked † are estimated (Nov 1975 surface temp not legible; April 1975 DO not measured). Lake Rotoiti max depth = 124 m; profiles shown to 50 m only — below 50 m temperature held at 50 m value; DO extrapolated linearly to near-zero in summer, held at 50 m value in winter (fully mixed).

hist_endpoints <- tibble::tribble(
  ~Date,          ~Surface_Temp, ~Deep_Temp, ~Surface_DO, ~Deep_DO,  ~thermo_depth, ~oxycline_depth,
  "1975-01-15",   24.0,  12.3,   9.1,   1.2,   10,   10,  
  "1975-02-16",   23.0,  12.4,  10.8,   0.5,   15,   15,
  "1975-03-16",   21.5,  13.5,  10.0,   0.6,   20,   20,  # temp Mar 16; DO Mar 14
  "1975-04-17",   18.0,  13.0,  NA,     NA,    25,   25,  # DO not measured
  "1975-05-19",   15.1,  12.5,  11.3,   0.7,   30,   30,  
  "1975-06-17",   12.6,  11.7,   9.0,   9.1,   NA,   NA,  # winter: uniform mixing
  "1975-07-14",   11.2,  10.5,  13.4,  11.0,   NA,   NA,
  "1975-08-05",   10.3,   9.9,  12.0,  11.6,   NA,   NA,
  "1975-09-15",   11.4,  10.6,  10.9,   7.5,   35,   35,
  "1975-10-15",   13.6,  11.1,  10.6,   8.2,   35,   35,
  "1975-11-10",   15.6,  11.3,  10.1,   6.5,   30,   30,
  "1975-12-08",   19.0,  12.1,   9.9,   6.0,   15,   15,
  # 1976
  "1976-01-13",   19.2,  11.9,   9.5,   4.8,   15,   15,
  "1976-02-09",   19.3,  12.0,  10.3,   1.2,   20,   20,  
  "1976-03-30",   18.7,  12.7,   9.2,   0.5,   20,   20,  
  "1976-04-20",   17.4,  13.0,   9.1,   1.2,   30,   30,  
  "1976-05-18",   14.4,  12.8,   9.2,   1.6,   40,   40,  
  "1976-06-23",   11.5,  10.6,   9.8,  10.3,   NA,   NA,
  "1976-07-20",   10.4,  10.4,  10.8,  10.7,   NA,   NA,
  "1976-08-17",   10.3,  10.1,  11.7,  10.7,   NA,   NA,
  "1976-09-22",   11.4,  10.4,  11.5,  10.7,   35,   35,
  "1976-10-19",   13.6,  10.9,  10.8,   9.1,   35,   35,
  "1976-11-18",   14.3,  11.2,  10.0,   7.0,   30,   30,
  "1976-12-14",   17.7,  11.4,   9.9,   5.2,   15,   15
) |>
  dplyr::mutate(
    Date      = as.Date(Date),
    Lake      = "Rotoiti",
    Month_num = lubridate::month(Date)
  )

# Generate depth profiles for each sampling date
max_depth_hist <- 124L   # Lake Rotoiti max depth

sigmoid_profile <- function(surface, deep, transition_d, depths, steepness = 3) {
  s <- 1 / (1 + exp(-(depths - transition_d) / steepness))
  surface + (deep - surface) * s
}

build_hist_profiles <- function(row) {
  depths <- 0:max_depth_hist

  # Temperature
  if (is.na(row$thermo_depth)) {
    temp_0_50 <- approx(c(0, 50), c(row$Surface_Temp, row$Deep_Temp), xout = pmin(depths, 50))$y
  } else {
    temp_0_50 <- sigmoid_profile(row$Surface_Temp, row$Deep_Temp, row$thermo_depth, pmin(depths, 50))
  }
  # Below 50 m: hold at 50 m temperature
  temp_vals <- ifelse(depths <= 50, temp_0_50, temp_0_50[51])

  # DO
  if (is.na(row$Surface_DO) || is.na(row$Deep_DO)) return(NULL)
  winter <- is.na(row$oxycline_depth)
  if (winter) {
    do_0_50 <- approx(c(0, 50), c(row$Surface_DO, row$Deep_DO), xout = pmin(depths, 50))$y
    # Below 50 m: hold at 50 m value in winter (lake is mixed)
    do_vals <- ifelse(depths <= 50, do_0_50, do_0_50[51])
  } else {
    do_0_50 <- sigmoid_profile(row$Surface_DO, row$Deep_DO, row$oxycline_depth, pmin(depths, 50), steepness = 5)
    # Below 50 m: extrapolate DO linearly to near zero (anoxic summer hypolimnion)
    do_50   <- do_0_50[51]
    do_vals <- ifelse(depths <= 50, do_0_50,
                      pmax(0, do_50 * (1 - (depths - 50) / (max_depth_hist - 50))))
  }

  tibble::tibble(Lake = row$Lake, Date = row$Date, Depth = depths, Temp = temp_vals, DO = do_vals)
}

hist_profiles_raw <- purrr::map(seq_len(nrow(hist_endpoints)), ~build_hist_profiles(hist_endpoints[.x, ]))
hist_profiles_raw <- dplyr::bind_rows(hist_profiles_raw[!sapply(hist_profiles_raw, is.null)])

# ── Profile verification plots ────────────────────────────────────────────────
hist_profiles_check <- hist_profiles_raw |>
  dplyr::mutate(
    Year      = lubridate::year(Date),
    Month_num = lubridate::month(Date)
  )

p_temp_check <- ggplot(
    hist_profiles_check |> dplyr::filter(Depth <= 50),
    aes(x = Temp, y = Depth, colour = Month_num, group = Date)
  ) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = TEMP_THRESHOLD, linetype = "dashed", colour = "red", linewidth = 0.5) +
  scale_y_reverse() +
  scale_colour_gradientn(colours = c("#2166ac", "#74add1", "#fee090", "#f46d43", "#d73027"),
    name = "Month", breaks = 1:12, labels = month.abb, limits = c(1, 12)
  ) +
  facet_wrap(~Year) +
  labs(x = "Temperature (°C)", y = "Depth (m)")

p_do_check <- ggplot(
    hist_profiles_check |> dplyr::filter(Depth <= 50),
    aes(x = DO, y = Depth, colour = Month_num, group = Date)
  ) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = DO_THRESHOLD, linetype = "dashed", colour = "red", linewidth = 0.5) +
  scale_y_reverse() +
  scale_colour_gradientn(
    colours = c("#2166ac", "#74add1", "#fee090", "#f46d43", "#d73027"),
    name = "Month", breaks = 1:12, labels = month.abb, limits = c(1, 12)
  ) +
  facet_wrap(~Year) +
  labs(x = "DO (mg/L)", y = "Depth (m)")

p_temp_check / p_do_check

# Join with Rotoiti hypsograph
rotoiti_hyps <- hypsograph_data1 |>
  dplyr::filter(grepl("Rotoiti", Lake, ignore.case = TRUE)) |>
  dplyr::mutate(Lake = "Rotoiti")

hist_profiles <- hist_profiles_raw |>
  dplyr::mutate(Depth = as.numeric(Depth)) |>
  dplyr::left_join(rotoiti_hyps |> dplyr::select(Lake, Depth, Area_at_Depth, Area_depth_0),
                   by = c("Lake", "Depth")) |>
  dplyr::filter(!is.na(Area_at_Depth)) |>
  dplyr::group_by(Lake, Date) |>
  dplyr::mutate(Cast_ID = dplyr::cur_group_id()) |>
  dplyr::ungroup()

livable_hist <- calc_livable_habitat_summary(
  df            = hist_profiles,
  source_label  = "Historical",
  lake_col      = "Lake",  date_col      = "Date",   cast_col      = "Cast_ID",
  depth_col     = "Depth", do_col        = "DO",     temp_col      = "Temp",
  area0_col     = "Area_depth_0",                    arealayer_col = "Area_at_Depth"
)

# Quick check plot
ggplot(livable_hist, aes(Date, Livable_Habitat_Percentage, colour = Season)) +
  geom_point(size = 2) +
  geom_line(aes(group = Season)) +
  labs(y = "Livable habitat (%)", x = NULL)
```

# Rotoiti — historical vs modern livable habitat
```{r}
#| label: fig-rotoiti-historical-combined
#| fig-width: 8
#| fig-height: 5

rotoiti_modern <- livable_habitat_all |>
  dplyr::filter(Lake == "Rotoiti")

rotoiti_combined <- dplyr::bind_rows(
  livable_hist    |> dplyr::mutate(Source = "Historical (1975–76)", Cast_ID = as.character(Cast_ID)),
  rotoiti_modern  |> dplyr::mutate(Source = as.character(Source),   Cast_ID = as.character(Cast_ID))
) |>
  dplyr::mutate(
    Source = factor(Source, levels = c("Historical (1975–76)", "Cast", "CTD")),
    Season = factor(Season, levels = season_order)
  )

rotoiti_seasonal <- rotoiti_combined |>
  dplyr::group_by(Year, Season, Source) |>
  dplyr::summarise(Mean_Livable = mean(Livable_Habitat_Percentage, na.rm = TRUE), .groups = "drop")

rotoiti_annual <- rotoiti_combined |>
  dplyr::group_by(Year, Source) |>
  dplyr::summarise(Annual_Mean = mean(Livable_Habitat_Percentage, na.rm = TRUE), .groups = "drop")

# Average annual mean across sources per year (single black line)
rotoiti_annual_avg <- rotoiti_annual |>
  dplyr::group_by(Year) |>
  dplyr::summarise(Annual_Mean = mean(Annual_Mean, na.rm = TRUE), .groups = "drop")

p_rotoiti_summary <- ggplot() +
  geom_point(data = rotoiti_seasonal, aes(Year, Mean_Livable, colour = Season, shape = Source), size = 2.5) +
  geom_line(data = rotoiti_seasonal, aes(Year, Mean_Livable, colour = Season, group = interaction(Season, Source), linetype = Source), linewidth = 0.7) +
  geom_line(data = rotoiti_annual_avg, aes(Year, Annual_Mean), colour = "black", linewidth = 1) +
  scale_shape_manual(values  = c("Historical (1975–76)" = 17, "Cast" = 16, "CTD" = 15)) +
  scale_linetype_manual(values = c("Historical (1975–76)" = "dashed", "Cast" = "solid", "CTD" = "dotted")) +
  labs(x = "Year", y = "Livable habitat (%)", colour = "Season", shape = "Source", linetype = "Source")

ggsave(file.path(out_dir, "fig-rotoiti-historical-combined.png"),
       p_rotoiti_summary, width = 8, height = 5, dpi = 300)

p_rotoiti_summary
```

# Rotoiti — long-term livable habitat trend (historical + modern)
```{r}
#| label: fig-rotoiti-trend
#| fig-width: 6
#| fig-height: 4

rotoiti_trend_fit  <- broom::tidy(lm(Annual_Mean ~ Year, data = rotoiti_annual_avg))
rotoiti_slope      <- rotoiti_trend_fit |> dplyr::filter(term == "Year")
rotoiti_slope_lbl  <- paste0(
  round(rotoiti_slope$estimate, 1), "% yr⁻¹ ",
  dplyr::case_when(
    rotoiti_slope$p.value < 0.001 ~ "***",
    rotoiti_slope$p.value < 0.01  ~ "**",
    rotoiti_slope$p.value < 0.05  ~ "*",
    TRUE                          ~ "ns"
  )
)

p_rotoiti_trend <- ggplot(rotoiti_annual_avg, aes(Year, Annual_Mean)) +
  geom_point(colour = "grey50", size = 2) +
  geom_smooth(method = "lm", se = TRUE, colour = "#2171b5", fill = "#2171b5", alpha = 0.15) +
  annotate("text", x = -Inf, y = Inf, label = rotoiti_slope_lbl,
           hjust = -0.15, vjust = 1.5, size = 3.5, colour = "grey30") +
  labs(x = "Year", y = "Annual mean livable habitat (%)")

ggsave(file.path(out_dir, "fig-rotoiti-trend.png"),
       p_rotoiti_trend, width = 6, height = 4, dpi = 300)

p_rotoiti_trend
```

# Trend analysis — is livable habitat changing over time?
```{r}
#| label: livable-trend
#| fig-width: 8
#| fig-height: 8
#| fig-cap: Long-term trends in livable habitat across Rotorua Te Arawa lakes (Cast + Buoy combined)

# Annual mean across Cast + Buoy CTD combined (pooled the same way as the summary figures)
annual_livable <- livable_habitat_all |>
  group_by(Lake, Year) |>
  summarise(Annual_Mean = mean(Livable_Habitat_Percentage, na.rm = TRUE), .groups = "drop")

livable_trends <- annual_livable |>
  group_by(Lake) |>
  group_modify(~ broom::tidy(lm(Annual_Mean ~ Year, data = .x))) |>
  filter(term == "Year") |>
  mutate(
    sig    = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                       p.value < 0.05 ~ "*", TRUE ~ "ns"),
    direction = if_else(estimate > 0, "increasing", "decreasing")
  ) |>
  select(Lake, estimate, std.error, p.value, sig, direction)

print(livable_trends)

trend_plot <- ggplot(annual_livable, aes(Year, Annual_Mean)) +
  geom_point(colour = "grey50", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "#2171b5", fill = "#2171b5", alpha = 0.15) +
  geom_text(
    data = livable_trends,
    aes(label = paste0(round(estimate, 1), "% yr⁻¹ ", sig)),
    x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3, colour = "grey30"
  ) +
  facet_wrap(~ Lake) +
  labs(x = "Year", y = "Annual mean livable habitat (%)")

ggsave(file.path(out_dir, "fig-livable-habitat-trends.png"), trend_plot, width = 8, height = 8, dpi = 300)

trend_plot
```

# Figures — livable habitat (Cast + Buoy combined)
```{r}
#| label: fig-livable-combined
#| fig-width: 8
#| fig-height: 8

# Time series: every cast/profile coloured by season
livable_habitat_plot_combined <- ggplot(
  livable_habitat_all,
  aes(Date, Livable_Habitat_Percentage, colour = Season)
) +
  geom_point(alpha = 0.6) +
  geom_line(aes(group = interaction(Season)), linewidth = 0.4, alpha = 0.6) +
  facet_wrap(~ Lake, ncol = 4) +
  labs(x = "Date", y = "Livable habitat (% of lake area)", colour = "Season")

livable_habitat_plot_combined
ggsave(file.path(out_dir, "fig-livable-habitat-combined.png"), livable_habitat_plot_combined,
       width = 8, height = 8, dpi = 300)

# Yearly/season summary with overall annual mean (black line)
livable_summary_all <- livable_habitat_all |>
  group_by(Lake, Year, Season) |>
  summarise(Mean_Livable_Habitat = mean(Livable_Habitat_Percentage, na.rm = TRUE), .groups = "drop")

yearly_mean_all <- livable_habitat_all |>
  group_by(Lake, Year) |>
  summarise(Yearly_Mean = mean(Livable_Habitat_Percentage, na.rm = TRUE), .groups = "drop")

livable_habitat_summary_plot_combined <- ggplot() +
  geom_point(data = livable_summary_all,
             aes(Year, Mean_Livable_Habitat, col = Season), size = 2) +
  geom_line(data = livable_summary_all,
            aes(Year, Mean_Livable_Habitat, col = Season, group = Season), linewidth = 0.8) +
  geom_line(data = yearly_mean_all,
            aes(Year, Yearly_Mean), col = "black", linewidth = 1) +
  facet_wrap(~ Lake) +
  labs(x = "Year", y = "Livable habitat (%)", col = "Season")

ggsave(file.path(out_dir, "fig-livable-habitat-summary.png"), livable_habitat_summary_plot_combined, width = 8, height = 8, dpi = 300)

livable_habitat_summary_plot_combined
```

# Figures — livable depth range (Cast + Buoy combined)
```{r}
#| label: fig-depth-combined
#| fig-width: 8
#| fig-height: 11

Depth_plot_combined <- depth_ranges_all |>
  ggplot(aes(x = Date)) +
  geom_linerange(aes(ymin = cast_shallowest, ymax = cast_deepest),
                 colour = "black", linewidth = 1) +
  geom_linerange(aes(ymin = shallowest_livable, ymax = deepest_livable, colour = Season),
                 linewidth = 0.7, na.rm = TRUE) +
  scale_y_reverse() +
  facet_wrap(~ Lake, scales = "free_y", ncol = 2) +
  labs(x = "Year", y = "Depth (m)")


ggsave(file.path(out_dir, "fig-depth-combined.png"), Depth_plot_combined, width = 11, height = 11, dpi = 300)
       
Depth_plot_combined
```

# Figures — livable depth as stacked area (alternative view)
```{r}
#| label: fig-depth-stacked-area
#| fig-width: 10
#| fig-height: 16

# Same depth_ranges_all data as Depth_plot_combined, but reduced to two stacked
# thicknesses (livable vs non-livable) per Date/Lake instead of absolute depth bands.
# Trades away *where* in the water column the livable zone sits for a compact view
# of *how much* of the water column is livable, over time.
depth_zones_long <- depth_ranges_all |>
  mutate(
    total_depth          = cast_deepest - cast_shallowest,
    livable_thickness     = pmax(dplyr::coalesce(deepest_livable, cast_shallowest) -
                                  dplyr::coalesce(shallowest_livable, cast_shallowest), 0),
    nonlivable_thickness = total_depth - livable_thickness
  ) |>
  select(Lake, Date, Season, livable_thickness, nonlivable_thickness) |>
  tidyr::pivot_longer(c(nonlivable_thickness, livable_thickness),
                       names_to = "Zone", values_to = "Thickness") |>
  mutate(Zone = factor(Zone,
                        levels = c("nonlivable_thickness", "livable_thickness"),
                        labels = c("Non-livable", "Livable habitat")))

Depth_area_plot_combined <- depth_zones_long |>
  ggplot(aes(x = Date, y = Thickness, fill = Zone)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = c("Non-livable" = "grey75", "Livable habitat" = "#1a9850")) +
  scale_y_reverse() +
  facet_wrap(~ Lake, scales = "free_y", ncol = 3) +
  labs(x = "Year", y = "Depth (m)", fill = NULL)

Depth_area_plot_combined
ggsave(file.path(out_dir, "fig-depth-stacked-area.png"), Depth_area_plot_combined,
       width = 10, height = 16, dpi = 300)
```


# Make int into lake maps
```{r}
#| label: lake-maps
#| fig-width: 8
#| fig-height: 11

# 2. HELPERS: PLOT ONE PERIOD (MONTH OR SEASON), PATCHWORK OF ALL LAKES

# Always show every lake we have bathymetry for, in the fixed display order, even when
# a given period has no cast for some of them (those just render as solid grey).
all_map_lakes <- lake_order[lake_order %in% names(bathy_rasters)]

# Lake shoreline outlines in NZTM x/y coordinates — overlaid on all pixel maps
# so each lake is always clearly delineated regardless of colour fill.
# all_lakes is created in the study-lakes-map chunk (EPSG:2193).
lake_boundaries <- purrr::map_dfr(all_map_lakes, function(lk) {
  lake_sf <- all_lakes |> dplyr::filter(lake_name == lk)
  if (nrow(lake_sf) == 0) return(NULL)
  coords <- sf::st_coordinates(sf::st_cast(sf::st_union(lake_sf), "MULTILINESTRING"))
  data.frame(x = coords[, "X"], y = coords[, "Y"], L = coords[, "L2"], Lake = lk)
})

# Nearest-depth match instead of exact equality, with one cast's classification per
# bathymetry point, then a majority vote across however many casts fall in this period
# (1 cast for a month, usually several for a season). A single cast rarely measures
# every exact integer depth present across the whole lake floor — especially the
# deepest basin pixels, which is why those showed up as blank holes under exact
# matching. Falls back to grey (NA) only if NONE of the period's casts had a
# measurement within max_dist_m of that point's depth — i.e. genuinely unmeasured.
match_nearest_depth_votes <- function(bathy_lake, profile_lake, cast_ids, max_dist_m = 3) {
  if (length(cast_ids) == 0) {
    return(bathy_lake |> mutate(Livable_Habitat = NA))
  }

  vote_matrix <- sapply(cast_ids, function(cid) {
    cast_depths <- profile_lake |> filter(Cast_ID == cid)
    if (nrow(cast_depths) == 0) return(rep(NA, nrow(bathy_lake)))
    idx <- vapply(bathy_lake$Depth, function(d) {
      diffs <- abs(cast_depths$Depth - d)
      i <- which.min(diffs)
      if (diffs[i] > max_dist_m) NA_integer_ else i
    }, integer(1))
    cast_depths$Livable_Habitat[idx]
  })
  if (is.null(dim(vote_matrix))) vote_matrix <- matrix(vote_matrix, ncol = length(cast_ids))

  majority <- apply(vote_matrix, 1, function(row) {
    row <- row[!is.na(row)]
    if (length(row) == 0) return(NA)
    mean(row) > 0.5
  })

  bathy_lake |> mutate(Livable_Habitat = majority)
}

plot_lake_period <- function(lake_name, cast_ids_this_period) {
  bathy_lake <- bathy_points_sub |>
    filter(Lake == lake_name) |>
    rename(Depth = depth)

  profile_lake <- profile_conditions |> filter(Lake == lake_name)

  df <- match_nearest_depth_votes(bathy_lake, profile_lake, cast_ids_this_period)

  ggplot(df, aes(x, y, col = Livable_Habitat)) +
    geom_point(size = 0.6) +
    scale_colour_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                        na.value = "grey80") +
    geom_path(
      data        = lake_boundaries |> dplyr::filter(Lake == lake_name),
      aes(x, y, group = L),
      colour      = "black", linewidth = 0.5, inherit.aes = FALSE
    ) +
    coord_equal() +
    labs(x = NULL, y = NULL, subtitle = lake_name) +
    theme(
      legend.position = "none",
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank()
    )
}

make_month_patchwork <- function(year_val, month_val) {
  month_cast_ids <- monthly_casts |>
    filter(Year == year_val, Month == month_val)

  title_label <- paste0("Livable habitat | ", as.character(month_val), " ", year_val)

  lake_plots <- lapply(all_map_lakes, function(lk) {
    cast_ids <- month_cast_ids |> filter(Lake == lk) |> pull(Cast_ID)
    plot_lake_period(lk, cast_ids)
  })

  wrap_plots(lake_plots, nrow = 4) +
    plot_annotation(title = title_label)
}

# Season-level equivalent — pools ALL casts within a Lake/Season/Season_Year via
# majority vote (see match_nearest_depth_votes above), rather than picking a single
# representative cast like the monthly version does.
seasonal_casts <- area_data_all_lakes |>
  distinct(Lake, Season_Year, Season, Cast_ID) |>
  arrange(Lake, Season_Year, Season)

make_season_patchwork <- function(season_year_val, season_val) {
  season_cast_ids <- seasonal_casts |>
    filter(Season_Year == season_year_val, Season == season_val)

  title_label <- paste0("Livable habitat | ", season_val, " ", season_year_val)

  lake_plots <- lapply(all_map_lakes, function(lk) {
    cast_ids <- season_cast_ids |> filter(Lake == lk) |> pull(Cast_ID)
    plot_lake_period(lk, cast_ids)
  })

  wrap_plots(lake_plots, nrow = 4) +
    plot_annotation(title = title_label)
}


# 3. PREVIEW: FIRST MONTH ONLY 

first_month_info <- monthly_casts |>
  arrange(Year, Month) |>
  slice(1)

first_year  <- first_month_info$Year
first_month <- first_month_info$Month

combined_first_month_plot <- make_month_patchwork(first_year, first_month)
combined_first_month_plot

# Season-level preview — same first period, but pooling every cast in that season
first_season_info <- seasonal_casts |>
  arrange(Season_Year, Season) |>
  slice(2380)

combined_first_season_plot <- make_season_patchwork(first_season_info$Season_Year, first_season_info$Season)
combined_first_season_plot


# 4. ALL Seasons → PNG FRAMES → VIDEO
# Disabled by default — slow (renders one plot per season) and not needed for every run.
# Set to TRUE to regenerate the frames + video.
if (FALSE) {

  # Directory to store frames
  frames_dir <- "livable_habitat_frames"
  if (!dir.exists(frames_dir)) dir.create(frames_dir, recursive = TRUE)

  # All Year–Month combinations present in seasonal_casts (sorted)
  ym_table <- seasonal_casts |>
    distinct(Year, Month) |>
    arrange(Year, Month)

  frame_files <- c()

  for (i in seq_len(nrow(ym_table))) {
    yr <- ym_table$Year[i]
    mn <- ym_table$Month[i]

    p <- make_month_patchwork(yr, mn)
    if (is.null(p)) next

    # safe filename, e.g. 2004-01_Rotorua_etc.png → here just year-month
    mn_num <- as.integer(mn)
    file_name <- sprintf("livable_habitat_%04d-%02d.png", yr, mn_num)
    file_path <- file.path(frames_dir, file_name)

    ggsave(filename = file_path, plot = p, width = 10, height = 6, dpi = 300)
    frame_files <- c(frame_files, file_path)
  }

  # Encode video from the PNG frames
  # Adjust framerate if you want slower/faster animation
  output_video <- file.path(frames_dir, "livable_habitat_video.mp4")
  av_encode_video(frame_files, output = output_video, framerate = 2)

}


```

# Per-pixel livable habitat trend maps
```{r}
#| label: pixel-trend-maps
#| fig-width: 10
#| fig-height: 14

# Unlike trend_plot (one scalar trend per lake), this fits a separate linear trend per
# bathymetry pixel — Livable (0/1) ~ Year, majority-voted across that year's casts —
# so it shows WHERE within a lake habitat is improving or declining, not just the
# lake-wide average. Cached to disk since it's expensive (one nearest-depth-vote pass
# per lake per year). Delete pixel_trend_maps.rds to force a recompute.
pixel_trend_rds <- file.path(der_data_dir, "pixel_trend_maps.rds")

if (file.exists(pixel_trend_rds)) {

  pixel_trends <- readRDS(pixel_trend_rds)

} else {

  compute_pixel_trend <- function(lake_name) {
    bathy_lake <- bathy_points_sub |>
      filter(Lake == lake_name) |>
      rename(Depth = depth)

    profile_lake <- profile_conditions |> filter(Lake == lake_name)
    years <- sort(unique(profile_lake$Year))
    if (length(years) < 3) return(NULL)  # not enough years for a meaningful trend

    # Pixel x Year matrix of livable classification (majority vote within each year,
    # reusing the same nearest-depth-vote logic as the season/month maps)
    year_matrix <- vapply(years, function(yr) {
      cast_ids <- profile_lake |> filter(Year == yr) |> pull(Cast_ID) |> unique()
      match_nearest_depth_votes(bathy_lake, profile_lake, cast_ids)$Livable_Habitat
    }, FUN.VALUE = logical(nrow(bathy_lake)))

    # Per-pixel OLS slope of (Livable as 0/1) ~ Year, skipping years with no data for
    # that specific pixel — needs at least 3 years of data to fit a trend at all.
    y_mat   <- matrix(as.numeric(year_matrix), nrow = nrow(year_matrix))
    n_valid <- rowSums(!is.na(y_mat))
    slope   <- rep(NA_real_, nrow(y_mat))

    for (i in which(n_valid >= 3)) {
      yi <- y_mat[i, ]
      ok <- !is.na(yi)
      if (length(unique(years[ok])) < 2) next
      slope[i] <- coef(lm(yi[ok] ~ years[ok]))[2]
    }

    bathy_lake |>
      mutate(Trend = slope, Lake = lake_name) |>
      filter(!is.na(Trend))
  }

  pixel_trends <- lapply(all_map_lakes, compute_pixel_trend) |> bind_rows()

  saveRDS(pixel_trends, pixel_trend_rds)

}

# Shared colour scale across all lakes so trend intensity is directly comparable —
# a dark red in one lake means the same thing as a dark red in another.
# A handful of extreme outlier pixels were stretching the colour scale so far that
# typical trend values all looked washed-out near white. Cap the scale at the 95th
# percentile of |Trend| instead of the true max, and squish anything beyond that to
# the most saturated colour rather than letting it go off-scale.
trend_cap <- quantile(abs(pixel_trends$Trend), 0.95, na.rm = TRUE)

plot_lake_trend <- function(lake_name) {
  df <- pixel_trends |> filter(Lake == lake_name)
  if (nrow(df) == 0) {
    return(ggplot() + theme_void() + labs(subtitle = lake_name))
  }
  ggplot(df, aes(x, y, col = Trend)) +
    geom_point(size = 0.6) +
    scale_colour_gradient2(low = "#d73027", mid = "grey90", high = "#2171b5",
                           midpoint = 0, limits = c(-trend_cap, trend_cap),
                           oob = scales::squish) +
    geom_path(
      data        = lake_boundaries |> dplyr::filter(Lake == lake_name),
      aes(x, y, group = L),
      colour      = "black", linewidth = 0.5, inherit.aes = FALSE
    ) +
    coord_equal() +
    labs(x = NULL, y = NULL, subtitle = lake_name, colour = "Trend") +
    theme(
      axis.text = element_blank(), axis.title = element_blank(),
      axis.ticks = element_blank(), axis.line = element_blank()
    )
}

trend_map_plots  <- lapply(all_map_lakes, plot_lake_trend)
combined_trend_map <- wrap_plots(trend_map_plots, nrow = 4) +
  plot_annotation(title = "Per-pixel trend in livable habitat (change in proportion of livable years, per year)")

combined_trend_map
ggsave(file.path(out_dir, "fig-pixel-trend-maps.png"), combined_trend_map,
       width = 10, height = 14, dpi = 300)
```

# Per-pixel livable habitat persistence maps
```{r}
#| label: pixel-persistence-maps
#| fig-width: 10
#| fig-height: 14

# Different question to the trend map: not "is this pixel changing?" but "how reliably
# is this pixel livable at all?" Built at the Season_Year x Season level (not collapsed
# to year first) so the per-season breakdown isn't confounded by within-year seasonal
# cycling — a pixel livable every winter and never in summer should NOT look the same
# as one that's moderately livable year-round. Cached to disk; delete
# pixel_persistence.rds to force a recompute.
pixel_persistence_rds <- file.path(der_data_dir, "pixel_persistence.rds")

if (file.exists(pixel_persistence_rds)) {

  pixel_persistence <- readRDS(pixel_persistence_rds)

} else {

  compute_pixel_persistence <- function(lake_name) {
    bathy_lake   <- bathy_points_sub |> filter(Lake == lake_name) |> rename(Depth = depth)
    profile_lake <- profile_conditions |> filter(Lake == lake_name)
    season_lake  <- seasonal_casts |> filter(Lake == lake_name)
    periods      <- season_lake |> distinct(Season_Year, Season)
    if (nrow(periods) < 3) return(NULL)

    # Pixel x Season-period matrix of livable classification (majority vote within
    # each individual season-year, same logic as the season maps)
    period_matrix <- vapply(seq_len(nrow(periods)), function(i) {
      cast_ids <- season_lake |>
        filter(Season_Year == periods$Season_Year[i], Season == periods$Season[i]) |>
        pull(Cast_ID)
      match_nearest_depth_votes(bathy_lake, profile_lake, cast_ids)$Livable_Habitat
    }, FUN.VALUE = logical(nrow(bathy_lake)))

    # Pooled persistence: proportion of ALL observed periods a pixel was livable
    pooled <- rowMeans(period_matrix, na.rm = TRUE)

    # Per-season persistence: proportion of that specific season's periods (e.g. every
    # observed summer) a pixel was livable
    by_season <- lapply(season_order, function(s) {
      cols <- periods$Season == s
      if (!any(cols)) return(rep(NA_real_, nrow(bathy_lake)))
      if (sum(cols) == 1) period_matrix[, cols] |> as.numeric()
      else rowMeans(period_matrix[, cols, drop = FALSE], na.rm = TRUE)
    })
    names(by_season) <- season_order

    bathy_lake |>
      mutate(Lake = lake_name, Pooled = pooled, !!!by_season) |>
      filter(!is.nan(Pooled))
  }

  pixel_persistence <- lapply(all_map_lakes, compute_pixel_persistence) |> bind_rows()

  saveRDS(pixel_persistence, pixel_persistence_rds)

}

plot_lake_persistence <- function(lake_name, value_col, title_label) {
  df <- pixel_persistence |> filter(Lake == lake_name)
  if (nrow(df) == 0 || all(is.na(df[[value_col]]))) {
    return(ggplot() + theme_void() + labs(subtitle = lake_name))
  }
  ggplot(df, aes(x, y, col = .data[[value_col]])) +
    geom_point(size = 0.6) +
    scale_colour_gradient(low = "grey90", high = "#00BFC4", limits = c(0, 1),
                          na.value = "grey95") +
    geom_path(
      data        = lake_boundaries |> dplyr::filter(Lake == lake_name),
      aes(x, y, group = L),
      colour      = "black", linewidth = 0.5, inherit.aes = FALSE
    ) +
    coord_equal() +
    labs(x = NULL, y = NULL, subtitle = lake_name, colour = title_label) +
    theme(
      axis.text = element_blank(), axis.title = element_blank(),
      axis.ticks = element_blank(), axis.line = element_blank()
    )
}

# Pooled — one map per lake, all seasons/years combined
pooled_persistence_plots <- lapply(all_map_lakes, plot_lake_persistence,
                                   value_col = "Pooled", title_label = "% livable")
combined_pooled_persistence <- wrap_plots(pooled_persistence_plots, nrow = 4) +
  plot_annotation(title = "Pixel persistence — proportion of all observed periods livable")

combined_pooled_persistence
ggsave(file.path(out_dir, "fig-pixel-persistence-pooled.png"), combined_pooled_persistence,
       width = 10, height = 14, dpi = 300)

# Per-season — four separate combined maps, one per season
season_persistence_maps <- lapply(season_order, function(s) {
  plots <- lapply(all_map_lakes, plot_lake_persistence, value_col = s, title_label = "% livable")
  combined <- wrap_plots(plots, nrow = 4) +
    plot_annotation(title = paste0("Pixel persistence — ", s, " (proportion of observed ", s, "s livable)"))
  ggsave(file.path(out_dir, paste0("fig-pixel-persistence-", tolower(s), ".png")), combined,
         width = 10, height = 14, dpi = 300)
  combined
})
names(season_persistence_maps) <- season_order

# Preview one season here; the others are saved to outputs/ but not all printed inline
season_persistence_maps[["Summer"]]
```


# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```